NLIZE: A Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models

### 2018-03-27 paper/background.tex

\subsection{Natural Language Inference}
\label{sec:languageInference}
Natural language inference (NLI)~\cite{DaganRothSammons2013} is an important machine understanding task in NLP.
The goal of NLI is to predict the relationship between a premise (\textbf{P}) sentence and a hypothesis (\textbf{H}) sentence. 
The prediction falls into one of three categories: \emph{entailment} (\textbf{E}), \emph{contradiction} (\textbf{C}), and \emph{neutral} (\textbf{N}).
A simple example is shown in Table.~\ref{table:NLI}.
In this case, the premise is ``Jim ate an apple''. 
The hypothesis statement ``Jim ate fruits'' can be concluded from the premise. Therefore, the relationship between the premise and hypothesis is \emph{entailment}. However, we should note that such a relationship is not necessarily reversible. Since the concept ``fruits'' is less restrictive than that of ``apple'', we cannot conclude ``Jim ate an apple'' from the statement ``Jim ate fruits''. The same logic applies to the hypothesis ``Jim ate a Fuji apple'', in which the premise neither implies nor opposes the hypothesis, and therefore, their relationship is \emph{neutral}.
%\emph{Contradiction} means the textual of hypothesis opposes that of premise;
%and \emph{Neutral} implies no conclusion can be made on either \emph{Entailment} or \emph{Contradiction}.

\begin{table}[htbp]
\label{table:NLI}
\centering
\caption{An illustration of natural language inference.}
 \begin{tabular}{c | c c c c} 
 \hline
  P / H & sentences & entail & contradict & neutral \\ [0.5ex] 
 \hline
 premise & Jim ate an apple. &  -  &  -  & - \\ 
 hypothesis & Jim ate fruits. & \checkmark &   &  \\
 hypothesis & Jim ate a banana. &  & \checkmark & \\
 hypothesis & Tom ate an apple. &  &  & \checkmark \\
 hypothesis & Jim ate a Fuji apple. &   &  & \checkmark \\
 %\hline
 %premise & Facebook's IPO electrified the general public. &  -  &  -  & - \\ 
 %hypothesis & Facebook went public. & \checkmark   &  &  \\
 %hypothesis & General Electric went public. &  &   & \checkmark \\
 %hypothesis & People ignored Facebook's IPO. &  & \checkmark & \\
 \hline
\end{tabular}
\end{table}

At first glance, the task of natural language inference may seem less practical compared to other well-known NLP challenges such as machine translation; however, the ability to distinguish entailment and contradiction relationship is fundamental to understanding natural language at large. 
%
Considering the ambiguousness of natural language and polysemy words, the inference task can become quite challenging (especially from the learning algorithm's point of view). Take the following sentences as an example (here \textbf{P} refers to premise, \textbf{H} refers to hypothesis):  (\textbf{P}) Facebook's IPO electrified the general public; (\textbf{H1}) Facebook went public; (\textbf{H2}) General Electric went public; (\textbf{H3}) People ignored Facebook's IPO. The literal similarity between ``Electric'' and ``electrified'' may trick the model to predict \textbf{H2} as \emph{entailment}. The model likely will also fail to understand the link between ``went public'' and ``IPO'', and therefore, mistake \textbf{H1} as \emph{neutral}.
%
%The grand challenge in Natural Language Processing (NLP) is to have machine to acquire
%deep comprehension of textual information. Major tasks include Natural Language Inference,
%Machine Translation, Questions Answering, Summarization, and etc.
%Recently, neural networks models have performed strongly on these tasks.
%
Recently, Bowman et al. introduced a large corpus~\cite{BowmanAngeliPotts2015} for the NLI tasks, which has helped spawn a new wave of effective neural network models for those tasks. In this work, we focus on the analysis of the decomposable attention model~\cite{parikh2016emnlp} on this dataset.

@@

\subsection{Neural Network Model in NLP}
%\shusen{What is the basic understanding of neural network What is the end-to-end}
There are many connections and distinctions between neural network models employed for vision and NLP tasks.
On the one hand, as discussed in the introduction, the discrete nature of words and sentences presents additional challenges for
interpreting the NLP model.
%
On the other hand, a majority of recent NLP neural networks share the nature of
end-to-end model, where the entire model operates as a black-box that takes
vectorized input and yields a final prediction for a specific task.
%
Many of the recent NLP end-to-end neural network models, despite having potentially drastically different network architectures, share a similar high-level design that consists of three distinct stages (encode, align, classify, see Fig.~\ref{fig:modelPipeline}). 
%
The existence of shared conceptual structures means the proposed tool, although designed for NLI task, can be generalized for a much broader set of applications.

@@

\subsection{Attention Mechanism}
\label{sec:attention}
Among the three stages, the second stage often constitutes the most crucial part of the model as it determines where the classifier will focus for generating a prediction. The operation to compute the alignment between words in the input is referred to as the attention mechanism. 
The introduction of the attention mechanism~\cite{bahdanau2014neural} allows
pairwise interaction between internal representations of words (\shusen{the original text is hidden states, isn't the pairwise interaction between the representation of words, are the hidden states refer to that?}). 
This interaction can be naturally explained as a form of alignment that exposes an interpretable layer in end-to-end neural networks.
%
%As the only interpretable part of the internal state of the network, understanding how attention information 
Recently, attention has contributed to many strongly performed NLP models
~\cite{parikh2016emnlp},~\cite{rush2015neural},~\cite{yang2016hierarchical},
~\cite{seo2016bidirectional},~\cite{schwartz2017high}.

\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=0.9\linewidth]{attentionIllustration}
 \caption{Attention can be naturally explained as a form of alignment that exposes an interpretable layer in end-to-end neural networks. On the left, the matrix shows the soft alignment between between the premise and hypothesis pair.}
\label{fig:attention}
\end{figure}

For the natural language inference task, as illustrated in Fig.~\ref{fig:attention}, the attention information can be represented as a matrix describing the soft alignment between words in the premise and the words in the hypothesis. As we can see in this example, the higher values indicate better alignments between words. The \emph{subject}, \emph{verb}, and \emph{object} between these two sentences are aligned correctly (words such as ``an'', which has less importance in determining the sentence relationship were assigned a lower score). Interestingly, the difference between the subject words does not affect their alignment. The classification stage can then utilize this information to determine if the subject of the sentence is different, and therefore, despite the other part of the sentences being identical, the relationship between the premise and hypothesis is neutral.

>>>>

\subsection{Natural Language Inference}
\label{sec:languageInference}
Natural Language Inference (NLI)~\cite{DaganRothSammons2013} is an important machine understanding task in NLP.
The goal of NLI is to predict the relationship between a premise (\textbf{P}) sentence and a hypothesis (\textbf{H}) sentence.
The prediction falls in one of three categories: \emph{entailment} (\textbf{E}), \emph{contradiction} (\textbf{C}), and \emph{neutral} (\textbf{N}).
A simple example is shown in Table~\ref{tab:NLI}.
In this case, the premise is ``Boy ate an apple''.
The hypothesis statement ``Kid ate fruits'' can be concluded from the premise. Therefore, the relationship between the premise and hypothesis is \emph{entailment}. However, we should note that such a relationship is not necessarily reversible. Since the concept ``fruit'' is less restrictive than that of ``apple'', we can not conclude ``Boy ate an apple'' from the statement ``Kid ate fruit''. The same logic applies to the hypothesis of ``Boy ate a Fuji apple.'', in which the premise neither imply nor oppose the hypothesis, therefore, their relationship is \emph{neutral}.
%\emph{Contradiction} means the textual of hypothesis opposes that of premise;
%and \emph{Neutral} implies no conclusion can be made on either \emph{Entailment} or \emph{Contradiction}.

\begin{table}[htbp]
\label{tab:NLI}
\centering
\caption{An illustration of natural language inference.}
 \begin{tabular}{c | c c c c}
 \hline
  P / H & sentences & entail & contradict & neutral \\ [0.5ex]
 \hline
 premise & Boy ate an apple. &  -  &  -  & - \\
 hypothesis & Kid ate fruits. & \checkmark &   &  \\
 hypothesis & Boy ate a banana. &  & \checkmark & \\
 hypothesis & Tom ate an apple. &  &  & \checkmark \\
 hypothesis & Boy ate a Fuji apple. &   &  & \checkmark \\
 %\hline
 %premise & Facebook's IPO electrified the general public. &  -  &  -  & - \\
 %hypothesis & Facebook went public. & \checkmark   &  &  \\
 %hypothesis & General Electric went public. &  &   & \checkmark \\
 %hypothesis & People ignored Facebook's IPO. &  & \checkmark & \\
 \hline
\end{tabular}
\end{table}

At the first glance, the task of natural language inference may seem less practical compared to other well-known NLP challenges such as machine translation; however, the ability to distinguish entailment and contradiction relationship is fundamental to understanding natural language at large.
%
Considering the ambiguousness of natural language and polysemy words, the inference task can become quite challenging (especially from learning algorithm's point of view). Take the following sentences as an example (here \textbf{P} refers to premise, \textbf{H} refers to hypothesis):  (\textbf{P}) Facebook's IPO electrified the general public; (\textbf{H1}) Facebook went public; (\textbf{H2}) General Electric went public; (\textbf{H3}) People ignored Facebook's IPO. The literal similarity between ``Electric'' and ``electrified'' may trick a model to predict \textbf{H2} as \emph{entailment}. A model likely will also fail to understand the link between ``went public'' and ``IPO'', therefore, mistake \textbf{H1} as \emph{neutral}.
%
%The grand challenge in Natural Language Processing (NLP) is to have machine to acquire
%deep comprehension of textual information. Major tasks include Natural Language Inference,
%Machine Translation, Questions Answering, Summarization, and etc.
%Recently, neural networks models have performed strongly on these tasks.
%
Recently, Bowman et al. introduced a large corpus~\cite{BowmanAngeliPotts2015} for the NLI tasks, which has helped spawn a new wave of effective neural network models for those tasks. In this work, we focus on the analysis of the decomposable attention model~\cite{parikh2016emnlp} on this dataset.

@@

\subsection{Neural Network Model in NLP}
%\shusen{What is the basic understanding of neural network What is the end-to-end}
There are many connections and distinctions between neural network models employed for vision and NLP tasks.
On the one hand, as discussed in the introduction, the discrete nature of words and sentences presents additional challenges for
interpreting the NLP model.
%
On the other hand, a majority of recent NLP neural networks share the nature of
end-to-end model, where the entire model operates as a black-box that takes
vectorized input and yields a final prediction for a specific task.
%

Many of the recent NLP end-to-end neural network models, despite having potentially drastically different network architectures, share a similar high-level design that consists of three distinct stages (encode, align, classify, see Fig.~\ref{fig:modelPipeline}).
%
The existence of shared conceptual structures means the proposed tool, although designed for NLI task, can be generalized for a much broader set of applications.

@@

\subsection{Attention Mechanism}
\label{sec:attention}
Among the three stages, the second stage often constitutes a crucial part of the model as it determines where the classifier will focus for generating a prediction. The operation to compute the alignment between words in the input is referred to as the attention mechanism. 
The introduction of the attention mechanism~\cite{bahdanau2014neural} allows
pairwise interaction between internal representations of words. 
This interaction can be naturally explained as a form of alignment that exposes an interpretable layer in end-to-end neural networks.
%
%As the only interpretable part of the internal state of the network, understanding how attention information 
Recently, attention has contributed to many strongly performed NLP models

~\cite{parikh2016emnlp},~\cite{rush2015neural},~\cite{yang2016hierarchical},
~\cite{seo2016bidirectional},~\cite{schwartz2017high}.

\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=0.9\linewidth]{attentionIllustration}

 \caption{Attention can be naturally explained as a form of alignment which exposes an interpretable layer in end-to-end neural networks. On the left, the matrix shows the attention between between premise and hypothesis pair. On the right, attention can also be interpreted as alignment.}
\label{fig:attention}
\end{figure}

For natural language inference task, as illustrated in Fig.~\ref{fig:attention}, the attention information can be represented as a matrix describing the soft alignment between words in the premise and the words in the hypothesis. As we can see in this example, the higher values indicate significant alignments between words. The \emph{subject}, \emph{verb}, and \emph{object} between these two sentences are aligned correctly (words, such as ``an'', which has less importance in determine sentence relationship was assigned a lower score). Interestingly, the difference between the subject words does not affect their alignment. The classification stage can then utilize this information to determine the subject of the sentence are different, therefore, despite the other part of the sentences are identical, the relationship between the premise and the hypothesis is neutral.

### 2018-03-27 paper/intro.tex
add figures

% (What is the high level motivation)
% What is the obsatcle for effectively apply machine learning model
With the recent advances in neural network based model, machine learning has gained unprecedented popularity, and its adaption permeated many fields of studies.
%
However, from researchers to practitioners, one often need to overcome many obstacles during the training, debugging, and tuning processes to unleash the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made are critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and come up with hypotheses for the cause of failure is the key to identify the limitation and improve upon existing models.
However, providing meaningful answers to these questions are incredibly challenging and often regarded as impossible by many.

Recently, sizable research has been developed to combat the model interpretability challenges. Most notably, the works~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018} on interpreting convolution neural network (CNN) model from both machine learning and visualization community have revealed the enormous potential and immediate benefits of model interpretation. At the same time, these works also remind us how little we truly understand about the inner mechanism of these deep neural networks.

Compare to image classification task, natural language processing (NLP) systems have innate restrictions (e.g., the discrete nature of words) that makes the road to interpretation more treacherous. For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) capture, can not be readily generalized to natural language. Image (pixel values) represents a continuous solution space, in which human can quickly recognize the existence (or the absence) of patterns or shapes. For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, in which the interpolation of two vectors (i.e., two words) do not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivate the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Moreover, despite many recent advances, the majority of the existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation analysis often leads to many \emph{``what if ...''} type of questions. Such as, what if we perturbed the current input? Will the prediction be stable? What if we change one of the critical internal states of the model, how would the modification affect the prediction? What if the current prediction is wrong, how and where could we apply minimal change to the model to produce the correct result? And how would it affect the internal state we care about? These type of queries form a natural way to gain understanding and come up with hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting. Hypothetically, we can code up specialized experiments for each of these scenarios. However, such a process is not only tedious but also ignores the iterative nature of the exploratory analysis. Often, these questions are not pre-determined. Instead, new exploration paths arise as we investigate and analyze previous observations.

In this work, we aim to provide immediate and informative answers to these \emph{``what if''} questions by combining the expressive power of visualization and direct online query/optimization of the neural network model. Instead of viewing the model as an invariant object, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the internal states of the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provides a new perspective to address the model interpretation problem.

We implement our vision of the perturbation-driven exploration in an interactive visualization system for natural language inference models~\cite{Parikh2016}.
However, the components of the visualization and the overall paradigm can be readily extended to other NLP tasks, such as question and answer, and text summarization, etc.
%
In the simplest form, the inference task asks the question about whether if we can infer the content of \textbf{sentence B} from \textbf{sentence A} (\emph{entail}), or are they contradict with each other (\emph{contradiction}), or are they talk about different/unrelated things (\emph{neutral}). Language inference addresses the fundamental challenge of measuring sentences' semantic similarity, and is one the core NLP tasks (see detail in Section~\ref{sec:languageInference}).

One recent advance in neural natural language process models is the introduction of \emph{attention mechanisms}~\cite{VaswaniShazeerParmar2017} (Section~\ref{sec:attention}). Intuitively, the attention helps capture which part of the input the model deemed more important for making a prediction. The attention is often represented as weights for individual words or pairs of words (i.e., the alignment between words in different sentences).
%
%\shusen{seems repeated some of the previous points}
There have been many speculations on how attention works in various models. By utilizing the proposed tool, the domain experts study how changes in sentence input, attention, or prediction affect each other, which help them develop deeper intuitive and alternative hypotheses. 
%we can perturb the attention and observer the change in the prediction, perturb the sentence input and observe the effects on the attention. And more importantly, we also allow the user to change prediction, where we apply a constrained optimization to find the least amount of parameter update to produce the user assign label. 
%Another word, 
%Such type of exploration is not only essential to interpret the language inference problem, but can also be generalized to other NLP model utilized the attention mechanisms.
Moreover, we enhanced the standard visual encoding (e.g., as a bipartite graph or as a matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar guided simplification of the visual representation.
%
Finally, as discussed in Section~\ref{sec:caseStudy}, the ability to examine how attention corresponds to the grammatical structure also enables domain experts to speculate the potential benefit of including the linguistic structure in the design of attention component of the model.

In summary, the key contributions of the proposed works is the following:
\begin{itemize}
  % \item Distill the visualization tasks for making sense of the role of attention in natural language inference (section~\ref{sec:task});
    \item Introduce the perturbation-driven exploration paradigm for exploring natural language processing models;

    \item Perform error analysis by examining stability of the predictions under perturbation;

    \item Enhance the visual encoding of the attention by overlaying sentence linguistic structure to allow grammar guided simplification;

    \item Enable the interrogation of the relationship between input sentences, model internals, and the prediction by interactively exploring how the change in one place affect other parts of the model.
\end{itemize}
%
% In addition,

>>>>

% (What is the high level motivation)
% What is the obstacle for effectively apply machine learning model
With the recent advances in neural network model, machine learning has gained unprecedented popularity, and its adaption has permeated many fields of studies.
%
However, from researchers to practitioners, one often need to overcome many obstacles during the training, debugging, and tuning processes to unleash the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made are critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and come up with hypotheses for the cause of the error is the key to identifying the limitation and improving upon existing models.
However, providing meaningful answers to these questions is incredibly challenging and often regarded as impossible by many.

Recently, sizable research has been developed to combat the model interpretability challenges. Most notably, the works~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018} on interpreting the convolution neural network (CNN) model from both the machine learning and visualization communities have revealed the enormous potential and immediate benefits of model interpretation. At the same time, these works also remind us how little we truly understand about the inner mechanism of these deep neural networks.

Compared to image classification task, natural language processing (NLP) systems have innate restrictions (e.g., the discrete nature of words) that make the road to interpretation more treacherous. For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) captures, cannot be readily generalized to natural language. An image (pixel values) represents a continuous solution space, in which human can quickly recognize the existence (or the absence) of patterns or shapes. For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, in which the interpolation of two vectors (i.e., two words) does not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivates the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Moreover, despite many recent advances, the majority of the existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation analysis often leads to many \emph{``what if ...''} type of questions. Such as, what if we perturbed the current input? Will the prediction be stable? What if we change one of the critical internal states of the model? How would the modification affect the prediction? What if the current prediction is wrong? How and where could we apply minimal change to the model to produce the correct result? And how would it affect the internal state we care about? These types of queries form a natural way to gain understanding and come up with hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting. Hypothetically, we can code specialized experiments for each of these scenarios. However, such a process is not only tedious but also ignores the iterative nature of the exploratory analysis. Often, these questions are not pre-determined. Instead, new exploration paths arise as we investigate and analyze previous observations.

In this work, we aim to provide immediate and informative answers to these \emph{``what if''} questions by combining the expressive power of visualization and direct online query/optimization of the neural network model. Instead of viewing the models as invariant objects, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the internal states of the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provides a new perspective to address the model interpretation problem.

We implement our vision of the perturbation-driven exploration in an interactive visualization system for natural language inference models~\cite{Parikh2016}.
However, the components of the visualization and the overall paradigm can be readily extended to other NLP tasks, such as question and answer, and text summarization, etc.
%
In the simplest form, the inference task asks the question about whether we can infer the content of \textbf{sentence B} from \textbf{sentence A} (\emph{entail}), or if they are contradict with each other (\emph{contradiction}), or whether two sentences talk about different/unrelated things (\emph{neutral}). Language inference addresses the fundamental challenge of measuring the semantic similarity of sentences, and is one the core NLP tasks (see detail in Section~\ref{sec:languageInference}).

One recent advance in neural natural language process models is the introduction of \emph{attention mechanisms}~\cite{VaswaniShazeerParmar2017} (Section~\ref{sec:attention}). Intuitively, the attention helps capture which part of the input the model is deemed more important for making a prediction. The attention is often represented as weights for individual words or pairs of words (i.e., the alignment between words in different sentences).
%
%\shusen{seems repeated some of the previous points}
There have been many speculations on how attention works in various models. By utilizing the proposed tool, the domain experts study how changes in sentence input, attention, or prediction affect each other, which helps them develop deeper intuitive and alternative hypotheses. 
%we can perturb the attention and observer the change in the prediction, perturb the sentence input and observe the effects on the attention. And more importantly, we also allow the user to change prediction, where we apply a constrained optimization to find the least amount of parameter update to produce the user assign label. 
%Another word, 
%Such type of exploration is not only essential to interpret the language inference problem, but can also be generalized to other NLP model utilized the attention mechanisms.
Moreover, we enhanced the standard visual encoding (e.g., as a bipartite graph or as a matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar-guided simplification of the visual representation.
%
Finally, as discussed in Section~\ref{sec:caseStudy}, the ability to examine how attention corresponds to the grammatical structure also enables domain experts to speculate about the potential benefits of including the linguistic structure in the design of the attention component of the model.

In summary, the key contributions of the proposed works are the following:
\begin{itemize}
  % \item Distill the visualization tasks for making sense of the role of attention in natural language inference (section~\ref{sec:task});
    \item Introduce the perturbation-driven exploration paradigm for exploring natural language processing models;

    \item Perform error analysis by examining the stability of the predictions under perturbation;

    \item Enhance the visual encoding of the attention by overlaying sentence linguistic structure to allow grammar-guided simplification;

    \item Enable the interrogation of the relationships among input sentences, model internals, and the prediction by interactively exploring how the change in one place affects other parts of the model.
\end{itemize}
%
% In addition,
